A generalization of the Bradley-Terry model for draws in chess

### 2019-05-14 inst/collusion_chess_draws.Rnw
work

Here I analyse chess games for draws using the hyper2 package.
}

@@

\cite{moul2009}: for instance, a player may have the choice of more
and less conservative strategies with different chances for a
draw.  Risk- averse players may pursue less aggressive strategies,
trading expected payout for reduced variance.

@@

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry
model~\citep{bradley1952} can be augmented to accommodate draws in a
number of ways.  Here I present a new augmentation of Bradley-Terry in
which a drawn chess match is regarded as a three-way competition
between the players and an entity that wins if the match is drawn.
The resulting likelihood functions are easily optimised numerically
and I present a number of results.

Consider first the standard Bradley-Terry model for $n$ players.
First we ignore the effect of playing white.  We wish to find numbers
(``strengths") $p_i$ for $1\leqslant i\leqslant n$ with $p_i\geqslant
0, \sum p_i=1$ such that the probability of $i$ beating $j\neq i$ is
$\frac{p_i}{p_i+p_j}$.  A suitable likelihood function, on the
assumption of independence of trials, would be

$$
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j}\right)^{w_{ij}}
\left(\frac{p_j}{p_i+p_j}\right)^{l_{ij}}
$$

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by playing white.  We
thus consider a chess match between two players to be a three-way
entity but with a friendly ghost with strength $p_W$ whose strength is
added to the white player.

>>>>

In inference problems where the dataset is outcomes of paired
comparisons, such as chess, the Bradley-Terry model can be augmented
to chess-specific considerations in a number of ways.  Here I present
a new generalization of Bradley-Terry in which a chess match is
regarded as a three-way competition between the two players and an
entity that wins if the match is drawn.  Bradley-Terry can be further
modified to account for the White player advantage in a number of
ways.  This methodology affords insight into players' strengths,
response to playing Black or White, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.}

@@

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be to two-way
competitive situations such as tennis in which one or other player is
idenfied as the winner, but the system has been applied to many
non-competitive situations such mating preference~\cite{hankin_aylmer}
and consumer behaviour.  The Bradley-Terry model~\citep{bradley1952}
considers $n$ players and assigns non-negative numbers
$p_1,\ldots,p_n$ with unit sum to each player.  Then player $i$ beats
player $j$ with probability $\frac{p_i}{p_i+p_j}$.  This simple system
has been modified and extended many times, for example to account for
multi-way choices such as icon selection~\cite{hankinhyperdirichlet}.

Bradley-Terry is not directly applicable to Chess for two reasons:
firstly, the White player is at a considerable
advantage~\cite{asdfasdffdsfds}; and secondly, many chess matches end
in a draw.  Considering White advantage first, we find an overall mean
of about $(+0.375=0.394-0.276)$.  However, interpretations of this
statistic vary and the issue is hotly debated with opinions ranging
from Rowson who compares first-move advantage to that of serving in
tennis, to that of commentators such as \citeauthor{adorjan} who
claims that White's advantage is a ``delusion''; and that Black can
mount an intelligent and dynamic response to white: ``chess is a game
of complete information, and Black's information is always
greater---by one move!''.

Chess games may be won by either player, or may be drawn. The
implication of a draw is that neither player is better than the other.
Before 1867, tournament games that were drawn were replayed and in
this context we note that this procedure reduces a chess match to a
Bernoulli trial, with probability one.  This system became unworkable
due to the large number of draws, resulting in the modern practice of
awarding each player of a drawn match half a point, rendering chess a
zero-sum game.  Drawn games are generally considered to be undesirable
as they are generally poorer in spectacular or striking elements such
as sharp tactical combinations or sacrifice.  Excessive drawing
dissuades sponsors, leaves spectators feeling cheated, and leaves
commentators with material to discuss with the public.
\cite{moul2009} concurs, conceptualising a player as being able to
adopt a strategy on a spectrum ranging from conservative to
aggressive, thus being able to control the probability of a draw.  He
suggests that risk-averse players may pursue less vigorous strategies,
trading expected payout for reduced variance.

such as chess, the Bradley-Terry model can
be augmented to chess-specific considerations in a number of ways.
Here I present a new generalization of Bradley-Terry in which a chess
match is regarded as a three-way competition between the two players
and an entity that wins if the match is drawn.  Bradley-Terry can be
further modified to account for the White player advantage in a number
of ways.  This methodology affords insight into players' strengths,
response to playing Black or White, and risk-aversion as manifested by
probability of drawing.  The likelihood functions arising are easily
optimised numerically.  I analyse a number of datasets of chess
results, including the infamous World Chess Championships 1963 in
which the Soviets were accused of collusion.  I conclude that the
scoreline of the Candidates Tournament (Cura\c{c}ao 1962) exhibits no
evidence of collusion, in agreement with previous work.  However, I
find strong evidence of increased Soviet proclivity to draw in the
Interzonal Tournament, Stockholm 1962.

Consider first the standard Bradley-Terry model for a finite number of
chess players.  First we ignore the effect of playing white.  We wish
to find numbers (``strengths") $p_i$ for $1\leqslant i\leqslant n$
with $p_i\geqslant 0, \sum p_i=1$ such that the probability of $i$
beating $j\neq i$ is $\frac{p_i}{p_i+p_j}$.  A suitable likelihood
function, on the assumption of independence of trials, would be

\begin{equation}
\prod_{i<j}
\left(\frac{p_i}{p_i+p_j}\right)^{w_{ij}}
\left(\frac{p_j}{p_i+p_j}\right)^{l_{ij}}
\end{equation}

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by playing white.  We
thus consider a chess match between two players to be a three-way
competition between the two ``real'' players and the draw monster, but
this time we interpret the white player as the combined strength of
the player himself and another entity, the White monster, whose
strength $p_W$ is added to the white player.
